Optimizing and Incrementalizing Collection Queries by AST Transformations

### 2017-07-25 Abstract.tex
Add \@ before sentence-ending periods likely to miss detection
Grepped with `[A-Z]\.`, so probably missed some.

To incrementalize higher-order programs by program transformation, we extend
\emph{finite differencing} \citep{Paige82FDC,Blakeley:1986:EUM,Gupta99MMV} and
develop the first approach to incrementalization by program transformation for
higher-order programs. Base programs are transformed to \emph{derivatives},
program that transform input \emph{changes} to output changes.
We prove that our incrementalization approach is correct: We develop the theory
underlying this approach for simply-typed and untyped $\lambda$-calculus, and
discuss extensions to System F.

>>>>

To incrementalize higher-order programs by program transformation, we extend
\emph{finite differencing} \citep{Paige82FDC,Blakeley:1986:EUM,Gupta99MMV} and
develop the first approach to incrementalization by program transformation for
higher-order programs. Base programs are transformed to \emph{derivatives},
program that transform input \emph{changes} to output changes.
We prove that our incrementalization approach is correct: We develop the theory
underlying this approach for simply-typed and untyped $\lambda$-calculus, and
discuss extensions to System F\@.

### 2017-07-25 aosd13/pepm13/caseStudyImpl.tex
Add \@ before sentence-ending periods likely to miss detection
Grepped with `[A-Z]\.`, so probably missed some.

\section{Lifting higher-order expressions}
\label{sec:lifting-ho}
We have shown how to lift first-order expressions; however, the interface of collections also uses higher-order methods, that is, methods that accept functions as parameters, and we need to lift them as well to reify the complete collection EDSL. For instance, the \code{map} method applies a function to each element of a collection. In this section, we describe how we reify such expressions of function type.

@@

Note how \code{map}'s parameter \code{f} has type \code{Exp[T] => Exp[U]} as necessary to enable Scala's syntax for anonymous functions and automatic lifting of the function body. This implementation would work for queries on lists, but does not support other collection types or queries where the collection type changes. We show in \cref{subsec:collections} how {\LoS} integrates with such advanced features of the Scala Collection EDSL.

@@

Note how the lifted query operations use \code{CanBuildFrom} to compute the same static return type as the corresponding non-lifted query operation would compute. This reuse of type-level machinery allows {\LoS} to provide the same interface as the Scala Collections EDSL.

@@

The subset of Scala we reify roughly corresponds to a typed lambda calculus with subtyping and type constructors. It does not include constructs for looping or recursion, so it should be strongly normalizing as long as application programmers do not add expression nodes with non-terminating interpretations. However, query writers can use the host language to create a reified expression of infinite size. This should not be an issue if {\LoS} is used as a drop-in replacement for the Scala Collection EDSL.

>>>>

\section{Lifting higher-order expressions}
\label{sec:lifting-ho}
We have shown how to lift first-order expressions; however, the interface of collections also uses higher-order methods, that is, methods that accept functions as parameters, and we need to lift them as well to reify the complete collection EDSL\@. For instance, the \code{map} method applies a function to each element of a collection. In this section, we describe how we reify such expressions of function type.

@@

Note how \code{map}'s parameter \code{f} has type \code{Exp[T] => Exp[U]} as necessary to enable Scala's syntax for anonymous functions and automatic lifting of the function body. This implementation would work for queries on lists, but does not support other collection types or queries where the collection type changes. We show in \cref{subsec:collections} how {\LoS} integrates with such advanced features of the Scala Collection EDSL\@.

@@

Note how the lifted query operations use \code{CanBuildFrom} to compute the same static return type as the corresponding non-lifted query operation would compute. This reuse of type-level machinery allows {\LoS} to provide the same interface as the Scala Collections EDSL\@.

@@

The subset of Scala we reify roughly corresponds to a typed lambda calculus with subtyping and type constructors. It does not include constructs for looping or recursion, so it should be strongly normalizing as long as application programmers do not add expression nodes with non-terminating interpretations. However, query writers can use the host language to create a reified expression of infinite size. This should not be an issue if {\LoS} is used as a drop-in replacement for the Scala Collection EDSL\@.

### 2017-07-25 chap-intro.tex
Add \@ before sentence-ending periods likely to miss detection
Grepped with `[A-Z]\.`, so probably missed some.

Moreover, many programming languages are far more expressive than RDBMSs.
Typical RDBMS can only manipulate SQL relations, that is multisets (or bags) of
tuples (or sometimes sets of tuples, after duplicate elimination). Typical
programming languages (PL) support also arbitrarily nested lists and maps of
data, and allow programmers to define new data types with few restrictions; a
typical PL will also allow a far richer set of operations than SQL.

@@

Furthermore, \cref{ch:cts} comes from a recently submitted
manuscript~\citep*{Giarrusso2018Static}. I
designed the overall approach, the transformation and the case study on
sequences and nested loops. Proofs were done in collaboration with Yann
Régis-Gianas: most of the actual mechanization work is his, though I contributed significantly to the
correctness proof for ILC. Philipp Schuster contributed to the evaluation,
devising language plugins for bags and maps in collaboration with me.

>>>>

Moreover, many programming languages are far more expressive than RDBMSs.
Typical RDBMS can only manipulate SQL relations, that is multisets (or bags) of
tuples (or sometimes sets of tuples, after duplicate elimination). Typical
programming languages (PL) support also arbitrarily nested lists and maps of
data, and allow programmers to define new data types with few restrictions; a
typical PL will also allow a far richer set of operations than SQL\@.

@@

Furthermore, \cref{ch:cts} comes from a recently submitted
manuscript~\citep*{Giarrusso2018Static}. I
designed the overall approach, the transformation and the case study on
sequences and nested loops. Proofs were done in collaboration with Yann
Régis-Gianas: most of the actual mechanization work is his, though I contributed significantly to the
correctness proof for ILC\@. Philipp Schuster contributed to the evaluation,
devising language plugins for bags and maps in collaboration with me.
